Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-20 paper/main.tex
another revamp of results related to fig. 6 following JRM/PCF discussion

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
question based on all other Quiz~1 questions, we observed an outcome consistent
with our null hypothesis: higher estimated knowledge at the embedding
coordinate of a held-out question was associated with a lower likelihood of
answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
26.545],\ p = 0.001$). This outcome suggests that our knowledge estimates do
\textit{not} provide useful information about participants' Quiz~1 performance
when we aggregated across all question content areas. We speculated that this
might either indicate that the knowledge estimates are uninformative in
general, or about Quiz~1 performance in particular. When we repeated this
analysis for Quizzes~2 and~3, the direction of this relationship reversed:
higher estimated knowledge for a given question predicted a greater likelihood
of answering it correctly (Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\
\textnormal{CI} = [14.966,\ 29.309],\ p = 0.002$; Quiz~3: $OR = 3.238,\
\lambda_{LR} = 6.882,\ 95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.017$).
Taken together, these results suggest that our knowledge estimates \textit{do}
reliably predict participants' likelihood of success on individual quiz
questions, but only when there is some ``contrast'' in performance between the
content of different questions. On Quiz~1, before participants have learned the
content of either lecture, their overall performance is poor (average
proportion correct: $0.46$), and the specific questions they answer correctly
not not seem to be structured in a meaningful way with respect to the embedding
space (e.g., Fig.~\ref{fig:knowledge-maps}A, left panel). On Quizzes~2 and~3,
after participants have begun their training, their overall performance
improves (average proportion correct for Quiz~2: 0.60; average proportion
correct for Quiz~3: 0.79), and the specific content areas they improve on are
captured by the lecture and question embeddings (and, therefore, our knowledge
estimates).

We observed a similar pattern of results when we fit GLMMs to estimates of
participants' knowledge for each question about one lecture derived from other
questions about the \textit{same} lecture. Specifically, for questions that
participants answered on Quiz~1, prior to watching either lecture, knowledge
for the embedding coordinates of \textit{Four Fundamental Forces}-related
questions estimated using other \textit{Four Fundamental Forces}-related
questions did not reliably predict whether those questions were answered
correctly ($OR = 1.891,\ \lambda_{LR} = 2.293,\ 95\%\ \textnormal{CI} =
[2.091,\ 2.622],\ p = 0.139$). The same was true of knowledge estimates for
\textit{Birth of Stars}-related questions based on other \textit{Birth of
Stars}-related questions ($OR = 0.722,\ \lambda_{LR} = 5.115,\ 95\%\
\textnormal{CI} = [0.094,\ 0.146],\ p = 0.738$). When we recomputed these
within-lecture knowledge estimates using questions from Quiz~2---which
participants took immediately after viewing \textit{Four Fundamental Forces}
but prior to viewing \textit{Birth of Stars}---we found that they now reliably
predicted success on \textit{Four Fundamental Forces}-related questions ($OR =
9.023,\ \lambda_{LR} = 18.707,\ 95\%\ \textnormal{CI} = [10.877,\ 22.222],\ p =
0.001$) but not on \textit{Birth of Stars}-related questions ($OR = 0.306,\
\lambda_{LR} = 5.115,\ 95\%\ \textnormal{CI} = [4.624,\ 5.655],\ p = 0.055$).
Using participants' responses from Quiz~3 (taken immediately after viewing
\textit{Birth of Stars}), we found that within-lecture knowledge estimates for
\textit{Birth of Stars}-related questions could now reliably predict success on
those questions ($OR = 5.467,\ \lambda_{LR} = 10.670,\ 95\%\ \textnormal{CI} =
[7.998, 12.532],\ p = 0.006$). In contrast, within-lecture knowledge estimates
for \textit{Four Fundamental Forces} questions answered on Quiz~3 were no
longer directly related to the likelihood of successfully answering them and
instead exhibited the inverse relationship we would expect to arise from
unstructured knowledge (with respect to the embedding space; $OR = 0.013,\
\lambda_{LR} = 14.648,\ 95\%\ \textnormal{CI} = [10.695, 23.096],\ p = 0.001$).
In principle, this ``prediction failure'' might arise from several possible
scenarios. One possibility is that the participants are answering nearly all of
the questions about \textit{Four Fundamental Forces} correctly on Quiz~3, and
(as with the Quiz~1 questions) there might be insufficient contrast in
performance across different questions to reliably estimate participants'
knowledge. However, when we examined participants' performance on Quiz~3
questions about \textit{Four Fundamental Forces}, we found that they answered
these questions correctly only 76\% of the time, on average (i.e., unlikely to
reflect a ``ceiling'' effect). Alternatively, we speculated that participants
might be forgetting some of the material from the \textit{Four Fundamental
Forces} lecture, and that this forgetting might be happening in a relatively
``random'' way with respect to spatial distance within the embedding space.
This might lead our approach to over-estimate knowledge for held-out questions
about ``forgotten'' knowledge that participants answered incorrectly. Indeed,
participants' performance on \textit{Four Fundamental Forces}-related questions
was slightly (thought not dramatically) lower on Quiz~3 than on Quiz~2 (mean
proportion correct on Quiz~2: $0.77$). Taken together, these results suggest
that our approach can distinguish between questions about different content
covered by a single lecture when participants have sufficiently structured
knowledge about its contents, though this specificity may decrease further in
time from when the lecture in question was viewed.

Finally, when we fit GLMMs to estimates of participants' knowledge for
questions about one lecture using questions they answered (on the same quiz)
about the \textit{other} lecture, we also observed a (largely) similar pattern.
On Quiz~1, we found that participants' abilities to correctly answer questions
about \textit{Four Fundamental Forces} could be predicted from their responses
to questions about \textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} =
7.205,\ 95\%\ \textnormal{CI} = [6.224, 7.524],\ p = 0.039$) and similarly,
that their ability to correctly answer \textit{Birth of Stars}-related
questions could be predicted from their responses to \textit{Four Fundamental
Forces}-related questions ($OR = 1.522,\ \lambda_{LR} = 6.448,\ 95\%\
\textnormal{CI} = [5.656, 6.843],\ p = 0.043$). This reflects coarse-scale
structure in participants' knowledge prior to any training in our experiment.
When we repeated this analysis using questions from Quiz~2, we found
participants' responses to \textit{Four Fundamental Forces}-related questions
did \textit{not} reliably predict their success on \textit{Birth of
Stars}-related questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\
\textnormal{CI} = [3.027, 3.600],\ p = 0.125$), nor did their responses to
\textit{Birth of Stars}-related questions reliably predict their success on
\textit{Four Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR}
= 3.266,\ 95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.094$). These
``prediction failures'' appear to come from the fact that any signal derived
from participants' knowledge about the content of the \textit{Birth of Stars}
lecture (prior to watching it) is swamped by the much more dramatic increase in
their knowledge about the content of the \textit{Four Fundamental Forces}
(which they watched just prior to taking Quiz~2). This is reflected in their
Quiz~2 performance for questions about each lecture (mean proportion correct
for \textit{Four Fundamental Forces}-related questions on Quiz~2: $0.77$; mean
proportion correct for \textit{Birth of Stars}-related questions on Quiz~2:
$0.36$). When we again carried out these across-lecture knowledge predictions
using questions from Quiz~3 (when participants had now viewed \textit{both}
lectures), we could again reliably predict success on questions about both
\textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\
\textnormal{CI} = [9.126, 18.476],\ p = 0.004$) and \textit{Birth of Stars}
($OR = 7.302,\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\
p = 0.017$) using responses to questions questions about the other lecture's
content. Across all three versions of these analyses, our results suggest that
(by and large) our knowledge estimations can reliably predict participants'
abilities to answer individual quiz questions, distinguish between questions
about similar content, and generalize across content areas, provided that
participants' quiz responses reflect a minimum level of ``real'' knowledge
about both content on which these predictions are based and that for which they
are made. Our results also indicate some potential limitations of our approach:
when the contrast in participants' knowledge within the embedding space is low
overall, our approach (often) cannot reliably predict their abilities to answer
individual quiz questions.

>>>>

Before presenting our results, it is worth considering three possible
explanations of why a participant might answer a given question correctly or
incorrectly. One possibility is that the participant simply \textit{guessed}
the answer. A second is that they selected an answer by mistake, despite
``knowing'' the correct answer. In both of these scenarios, the participant's
knowledge about the question's content should be uninformative about their
observed response. A third possibility is that the participant's response
reflects their \textit{actual} knowledge about the question's content. In this
case, we \textit{might} expect to see a positive relationship between the
participant's knowledge and their likelihood of answering the question
correctly. However, in order to see this positive relationship, the
participant's knowledge must be structured in a way that is reflected (at least
partially) by the embedding space. In other words, if the participant's
performance reflects their true kwowledge, but our text embedding space does
not sufficiently capture the structure of that knowledge, then the knowledge
estimates we generate will not be predictive of the participant's performance.
In the extreme, if the embedding space is completely unstructured with respect
to the content of the quiz questions, then we would expect to see the negative
relationship between estimated knowledge and performance that we described
above.

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
question based on all other Quiz~1 questions, we observed an outcome consistent
with our null hypothesis: higher estimated knowledge at the embedding
coordinate of a held-out question was associated with a lower likelihood of
answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
26.545],\ p = 0.001$). This outcome suggests that our knowledge estimates do
\textit{not} provide useful information about participants' Quiz~1 performance
when we aggregated across all question content areas. We speculated that this
might either indicate that the knowledge estimates are uninformative in
general, or about Quiz~1 performance in particular. This would be expected, for
example, if participants were guessing about the answers to the Quiz~1
questions (prior to having watched either lecture). When we repeated this
analysis for Quizzes~2 and~3, we found that \textit{higher} estimated knowledge
for a given question predicted a greater likelihood of answering it correctly
(Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\ \textnormal{CI} =
[14.966,\ 29.309],\ p = 0.002$; Quiz~3: $OR = 3.238,\ \lambda_{LR} = 6.882,\
95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.017$). Taken together, these
results suggest that our knowledge estimates reliably predict participants'
performance on individual held-out quiz questions, but only after participants
have received at least some training.

We observed a similar pattern of results when used this approach to estimate
participants' knowledge about held-out questions from one lecture using their
performance on other questions from the \textit{same} lecture. Specifically,
for Quiz~1 questions (i.e., prior to watching either), participants' estimated
knowledge for the embedding coordinates of held-out \textit{Four Fundamental
Forces}-related questions estimated using other \textit{Four Fundamental
Forces}-related questions did not reliably predict whether those questions were
answered correctly ($OR = 1.891,\ \lambda_{LR} = 2.293,\ 95\%\ \textnormal{CI}
= [2.091,\ 2.622],\ p = 0.139$). The same was true of knowledge estimates for
held-out \textit{Birth of Stars}-related questions based on other \textit{Birth
of Stars}-related questions ($OR = 0.722,\ \lambda_{LR} = 5.115,\ 95\%\
\textnormal{CI} = [0.094,\ 0.146],\ p = 0.738$). As in our analysis that
included all questions, we speculate that these ``null'' results might reflect
some degree of random guessing on Quiz~1. When we repeated these within-lecture
analyses using questions from Quiz~2 (which participants took immediately after
viewing \textit{Four Fundamental Forces} but prior to viewing \textit{Birth of
Stars}), we found that they now reliably predicted success on \textit{Four
Fundamental Forces}-related questions ($OR = 9.023,\ \lambda_{LR} = 18.707,\
95\%\ \textnormal{CI} = [10.877,\ 22.222],\ p = 0.001$) but not on
\textit{Birth of Stars}-related questions ($OR = 0.306,\ \lambda_{LR} = 5.115,\
95\%\ \textnormal{CI} = [4.624,\ 5.655],\ p = 0.055$). Here, we speculate that
participants might have been guessing about the \textit{Birth of Stars} content
(e.g., prior to having watched it), whereas they might have been drawing on
some structured knowledge about the \textit{Four Fundamental Forces} content
(e.g., from having just watched it). When we applied this approach to Quiz~3
responses (given immediately after viewing \textit{Birth of Stars}), we found
that within-lecture knowledge estimates for \textit{Birth of Stars}-related
questions could now reliably predict success on those questions ($OR = 5.467,\
\lambda_{LR} = 10.670,\ 95\%\ \textnormal{CI} = [7.998, 12.532],\ p = 0.006$).
However, within-lecture knowledge estimates for \textit{Four Fundamental
Forces} questions answered on Quiz~3 were no longer directly related to the
likelihood of successfully answering them and instead exhibited the inverse
relationship we would expect to arise from unstructured knowledge (with respect
to the embedding space; $OR = 0.013,\ \lambda_{LR} = 14.648,\ 95\%\
\textnormal{CI} = [10.695, 23.096],\ p = 0.001$). Speculatively, we suggest
that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content (e.g., perhaps in favor of prioritizing encoding
the just-watched \textit{Birth of Stars} content in preparation for the third
quiz). If this forgetting happens in a relatively ``random'' way (with respect
to spatial distance within the embedding space), then it could explain why some
held-out questions about \textit{Four Fundamental Forces} were answered
incorrectly, even if questions at nearby coordinates (i.e., about similar
content) were answered correctly. This might lead our approach to over-estimate
knowledge for held-out questions about ``forgotten'' knowledge that
participants answered incorrectly. Taken together, these within-lecture results
suggest that our approach can distinguish between questions about different
content covered by a single lecture when participants have sufficiently
structured knowledge about its contents, though this specificity may decrease
with time since the relevant material was learned.

Finally, we used this approach to estimate participants' knowledge about
held-out questions from one lecture using their performance on other questions
from the \textit{other} lecture. Here we again observed a similar pattern of
results, though with some notable differences. On Quiz~1, we found that
participants' abilities to correctly answer questions about \textit{Four
Fundamental Forces} could be predicted from their responses to questions about
\textit{Birth of Stars} ($OR = 1.896,\ \lambda_{LR} = 7.205,\ 95\%\
\textnormal{CI} = [6.224, 7.524],\ p = 0.039$) and similarly, that their
ability to correctly answer \textit{Birth of Stars}-related questions could be
predicted from their responses to \textit{Four Fundamental Forces}-related
questions ($OR = 1.522,\ \lambda_{LR} = 6.448,\ 95\%\ \textnormal{CI} = [5.656,
6.843],\ p = 0.043$). Given the results from our analyses that included all
questions and within-lecture predictions, we were surprised to find that the
knowledge estimates could reliably (if weakly) predict participants'
performance across content from different lectures. It is possible that this
result reflects a combination of random guessing prior to training (leading to
a weak effect size), in combination with some coarse-scale structured knowledge
that participants had about the content prior to watching either lecture. When
we repeated this analysis using questions from Quiz~2, we found participants'
responses to \textit{Four Fundamental Forces}-related questions did
\textit{not} reliably predict their success on \textit{Birth of Stars}-related
questions ($OR = 1.865,\ \lambda_{LR} = 3.205,\ 95\%\ \textnormal{CI} = [3.027,
3.600],\ p = 0.125$), nor did their responses to \textit{Birth of
Stars}-related questions reliably predict their success on \textit{Four
Fundamental Forces}-related questions ($OR = 3.490,\ \lambda_{LR} = 3.266,\
95\%\ \textnormal{CI} = [3.033, 3.866],\ p = 0.094$). These ``prediction
failures'' appear to come from the fact that any signal derived from
participants' knowledge about the content of the \textit{Birth of Stars}
lecture (prior to watching it) is swamped by the much more dramatic increase in
their knowledge about the content of the \textit{Four Fundamental Forces}
(which they watched just prior to taking Quiz~2). This is reflected in their
Quiz~2 performance for questions about each lecture (mean proportion correct
for \textit{Four Fundamental Forces}-related questions on Quiz~2: $0.77$; mean
proportion correct for \textit{Birth of Stars}-related questions on Quiz~2:
$0.36$). When we again carried out these across-lecture knowledge predictions
using questions from Quiz~3 (when participants had now viewed \textit{both}
lectures), we could again reliably predict success on questions about both
\textit{Four Fundamental Forces} ($OR = 11.294),\ \lambda_{LR} = 11.055,\ 95\%\
\textnormal{CI} = [9.126, 18.476],\ p = 0.004$) and \textit{Birth of Stars}
($OR = 7.302,\ \lambda_{LR} = 7.068,\ 95\%\ \textnormal{CI} = [6.490, 8.584],\
p = 0.017$) using responses to questions questions about the other lecture's
content. Across all three versions of these analyses, our results suggest that
(by and large) our knowledge estimations can reliably predict participants'
abilities to answer individual quiz questions, distinguish between questions
about similar content, and generalize across content areas, provided that
participants' quiz responses reflect a minimum level of ``real'' knowledge
about both content on which these predictions are based and that for which they
are made. Our results also indicate some important limitations of our approach:
if participants' quiz performance does not reflect what they know (e.g., when
they ``guess''), or if their knowledge is not structured in a way that is
reflected by the embedding space, then our knowledge estimates will not be
predictive of their performance.
